Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2021-02-04 paper/main.tex
paper s45

An equivalent definition for the Kendall's-$\tau$ distance counts the number of adjacent swaps that have to be made to convert $\sigma_1^{-1}$ into $\sigma_2^{-1}$ and, therefore, it is sometimes called \emph{swap distance}~\citep{ZaeStoBar2014:ppsn}. \EKHINE{Mirandolo bien, lo que pasa es que dicen que es lo mismo la swap distance que la kendall!}

@@

As shown in Algorithm~\ref{alg:umm}, UMM  starts with a \emph{small} sample of $\minit$ permutations that are generated uniformly at random  (although it would be possible to employ Latin Hyper-Cube sampling or other methods) and evaluated. Then, using the current sample $S$, the uBorda algorithm is applied to learn a reference permutation $\hat{\sigma}_0$ (line~\ref{line:uborda}). 
Together with the parameter $\theta$, they define the Mallows model $MM(\hat\sigma_0, \theta)$. 
A decreasing $\theta$ parameter (line~\ref{line:theta}) implies a decreasing variance of the Mallows model (Eq.~\ref{eq:MM}), as explained in  Section~\ref{sec:sampling} below. 
Then, we sample a new candidate permutation from  $MM(\hat\sigma_0, \theta)$ (line~\ref{line:sample}). This candidate permutation (solution) is evaluated and added to the sample $S$. The process is repeated up to a given maximum budget of evaluations $\FEmax$.
\EKHINE{(3) In section 3, second half of the second paragraph (from "Then, we sample a new candidate [...]"): the algorithm 1 is not properly described. Please provide a description in line with algorithm 1 and, above all, describe the rationale behind the choice of line 7 of algorithm 1.
I like it, but it deserves to be better described. It seems to me a sort of "novelty-driven" optimization.(4) A small detail, but requires a clarification. In page 6, learning step paragraph: the weight formula uses fitness in the exponent, while the base rho is in (0,1). Clearly the authors are working with minimization problems and want to increase the weight when fitness improves/decreases. This is ok, but this only happens with their formula when fitnesses are greater than 1. I know this is quite usual in the problems analyzed by the authors, but it may not be so usual in general. Please add a clarification.}

The key idea behind UMM is that the probabilistic Mallows model (MM) used in
UMM does not simply represent the current sample $S$. Instead, the model
assigns higher probability values to the permutations with known (or expected)
minimal fitness values. This is achieved by means of the uBorda ranking
(Section~\ref{sec:uborda}) with $w(\sigma) = \rho^{f(\sigma)}$ and $\rho \in (0,1)$, which %  can be seen as replicating each permutation
% $\sigma \in S$ a number of times inversely proportional to its fitness
% $f(\sigma)$, i.e., it 
behaves as if there were many copies of the best
solutions and few of the bad ones. A property of UMM is that the
computational complexity of working with the original sample (Borda) and with the
transformed sample (uBorda) is the same, i.e., polynomial. The most important steps are the update
(\emph{learning}) of the parameters $\sigma_0$, $\theta$ and $\rho$ (line~\ref{line:uborda} in Algorithm~\ref{alg:umm}) and
the sampling from $MM(\sigma_0,\theta)$ (line~\ref{line:sample}) %(lines~\ref{line:sample}--\ref{line:distance}). % in Algorithm~\ref{alg:umm}).

@@

\subsection{Learning step}
%
At each iteration, we start by fitting a MM to the sample $S$, which implies
estimating $\hat\sigma_0$ and $\theta$. Parameter $\theta$ is updated
deterministically at the sampling step.  The parameter $\sigma_0$, however, is
learned from the current sample $S$ by using the uBorda algorithm with a weight
$w(\sigma)=\rho^{f(\sigma)}$ (Section~\ref{sec:uborda}). The learning rate is
controlled by parameter $0 < \rho < 1$. \MANUEL{Don't we have to say something about $f(\sigma) > 1$?}
In UMM, the value of $\rho$ depends on the set of true fitness evaluations $F$
and it is set at each iteration such that the largest $100r_1$\% of the mass of
the weights is concentrated in the best $100r_2$\% of the solutions in $S$:
%
\MANUEL{Reviewer 3:  Justify the form of the formula used to learn the "learning rate" $\rho$ (equation 4). Why this particular approach was chosen?}
% 
\begin{equation}
%  \begin{split}
    r_1 \cdot \sum_{\tuple{\sigma,f(\sigma)}\in S}\rho^{f(\sigma)} =  \sum_{\tuple{\sigma', f(\sigma')}\in S'}\rho^{f(\sigma')}  %\\
  %    & \text{where}\;|S'| = r_2|S| \land \forall \sigma'\in S', \sigma\in S\setminus S' : f(\sigma')\leq f(\sigma).
%  \end{split}
\end{equation}
%
where $|S'| = r_2|S| \land \forall \tuple{\sigma',f(\sigma')}\in S', \tuple{\sigma,f(\sigma)}\in S\setminus S' : f(\sigma')\leq f(\sigma)$, that is, $S'$ contains the $r_2|S|$ elements of $S$ with the best  fitness values. %The values of $r_1, r_2 \in (0, 1)$ are user-defined and control the balance between exploration and intensification. %: extreme different values bias the sampling towards the best solutions found so far, which may lead to quick stagnation, whereas values (0.5,0.5) make the sampling more uniform, thus slowing down the learning.

@@

\subsection{Convergence guarantees}
\ekhine{}{completar}

@@

\subsubsection{UMM}
%
UMM uses the same budget of evaluations ($\FEmax$) as CEGO and initially
samples $\minit=10$ solutions uniformly.
%At each sampling step, we set
%$\beta=10$, meaning that 10 permutations are generated and the farthest from
%$S$ is selected for evaluation.
The other two parameters of UMM are $r_1$ and $r_2$, which determine the
balance between exploration and exploitation. In the experimental section we
will study their impact.

@@

We use the original implementation of GECO.\footnote{\url{https://cran.r-project.org/package=CEGO}} %
Although it is never stated in the original paper, the implementation of CEGO
generates a set of initial solutions of size $\minit=10$ by means of a
max-min-distance sequential design: new solutions are added to the set
sequentially by maximizing the minimum distance to solutions already in the
set. These initial solutions are then evaluated on the true fitness function
and the result is used to build the initial surrogate
model. % \citet{ZaeStoBar2014:ppsn} uses \minit=10
Following the authors of
CEGO~\citep{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn}, we use a GA to optimize
the surrogate models with population size of 20, crossover rate of 0.5,
mutation rate of $1/n$, tournament selection of size 2 with probability of 0.9,
interchange mutation (i.e., exchanging two randomly selected elements) and
cycle crossover for permutations. To generate each new solution, CEGO runs the
GA using the model as a surrogate of the true fitness function.  The authors of
CEGO used a budget of $10^4$ surrogate-model evaluations for each run of the
GA. However, with such a setting, a single run of CEGO with a budget of
$\FEmax=400$ true fitness evaluations requires \emph{more than a week} of
computation time, which was infeasible for our study. Due to extremely long
runtimes of CEGO and constraints in our computing system, we set the budget of
the GA to $10^3$ and we stopped each run of CEGO after 144 hours of CPU-time.
Although a higher GA budget could improve the results of CEGO reported here,
our results are similar (and sometimes better) than the ones reported in the
original papers~\citep{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn} due to
running CEGO with $\FEmax=400$ instead of $\FEmax=200$. We found that
doubling $\FEmax$ produces a smaller increase in runtime than multiplying the
GA budget by a factor of $10$. Nevertheless, we caution that choosing a
different trade-off may lead to different conclusions and more work is needed
to understand CEGO parameters.

\subsubsection{Other settings}
%
For simplicity, we use Kendall's-$\tau$
distance~(Eq.~\ref{eq:kendall_decomp}) in all experiments.
%Nevertheless,
%\citet{ZaeStoFriFisNauBar2014} points out that CEGO performs best on the QAP
%when using Hamming distance and the same may apply to UMM.
We plan to extend UMM to other distance measures in the future.  In all
experiments, we consider a maximum budget of $\FEmax=400$ evaluations of the
actual objective function. In a white-box context, state-of-the-art algorithms
for the LOP and PFSP typically evaluate thousands of solutions,
thus, the budget considered here for the black-box context is extremely
limited.

\subsubsection{Benchmark problems}
%
Experiments with real-world expensive black-box problems would be computationally infeasible. Instead, we consider classical combinatorial optimization problems as black-box optimization problems. For consistency with the rest of the paper, a permutation $\sigma$ denotes a ranking also in this section, thus we use the inverse permutation $\sigma^{-1}$ in the definition of the problems. Later in the experimental section we discuss the effects of not performing this inversion.
%
The \emph{Linear Ordering Problem (LOP)}, when minimizing, is defined as
%\footnote{For consistency with the other problems, we minimise the lower triangle instead of maximising the upper triangle.} 
\begin{equation}\label{eq:lop}
  \min_{\sigma \in S_n} f(\sigma) = \sum_{i=1}^{n} \sum_{j=1}^{i-1}  a_{\sigma^{-1}(i), \sigma^{-1}(j)}
\end{equation}
%
where $[a_{i,j}]$ is a matrix of size $n \times n$.

@@

The \emph{Permutation Flowshop Scheduling Problem (PFSP)} is defined by a matrix $[p_{ij}]$ of size $n \times M$ that gives the processing time of a job $i \in [n]$ on a machine $j \in [M]$. All jobs must be processed by all machines in the same order. Given a permutation of the jobs $\sigma$, $C_{i,j}$ denotes the completion time of job at position $\sigma^{-1}(i)$ on machine $j$, and $C_{\max} = C_{n,M}$  is the completion time of the last job on the last machine, i.e., the makespan. The objective of the PFSP is to find:
\begin{equation}\label{eq:pfsp}
  \begin{split}
    \min_{\sigma \in S_n}\quad& f(\sigma) = C_{n,M}\\
    \text{s.t.}\quad & C_{1,j} = 0 \quad j\in [M], \qquad C_{i,1} = 0\quad i \in [n]\\
    &C_{i,j} = p_{\sigma^{-1}(i),j} + \max\{C_{i-1, j}, C_{i, j-1}\}  \ \  i \in \{2,\dotsc,n\},\\
    &\hspace{14.5em} j \in\{2,\dotsc, M\}
  \end{split}
\end{equation}

@@

\subsubsection{Computing environment.}
% Due to the computational effort required, we used several computing systems to
% run experiments.
% All UMM experiments were either on run on Intel Xeon
% \emph{Broadwell} E5-2683v4 CPUs at 2.10\,GHz or Intel Xeon \emph{Haswell} E5-2680v3
% CPUs at 2.50\,GHz, both of them with 32\,GB RAM. 
 %
Experiments were run on Intel Xeon \emph{Ivybridge} E5-2650v2 CPUs at
2.60\,GHz, 64\,GB RAM running CentOS Linux release 7 (Core).
%The operating system was CentOS Linux release 7 (Core).

>>>>

An equivalent definition for the Kendall's-$\tau$ distance counts the number of adjacent swaps that have to be made to convert $\sigma_1^{-1}$ into $\sigma_2^{-1}$ and, therefore, it is sometimes called \emph{swap distance}~\citep{ZaeStoBar2014:ppsn}. %\EKHINE{Mirandolo bien, lo que pasa es que dicen que es lo mismo la swap distance que la kendall!}

@@

As shown in Algorithm~\ref{alg:umm}, UMM  starts with a \emph{small} sample of $\minit$ permutations that are generated uniformly at random  (although it would be possible to employ Latin Hyper-Cube sampling or other methods) and evaluated. Then, using the current sample $S$, the uBorda algorithm is applied to learn a reference permutation $\hat{\sigma}_0$ (line~\ref{line:uborda}). 
Together with the parameter $\theta$, the algorithm defines the Mallows model $MM(\hat\sigma_0, \theta)$. 
A decreasing $\theta$ parameter along iterations (line~\ref{line:theta}) implies a decreasing variance of the Mallows model (Eq.~\ref{eq:MM}), as explained in  Section~\ref{sec:sampling} below. 
Then, we sample a new candidate permutation from  $MM(\hat\sigma_0, \theta)$ (line~\ref{line:sample}). This candidate permutation (solution) is evaluated and added to the sample $S$. The process is repeated up to a given maximum budget of evaluations $\FEmax$.
\EKHINE{(3) In section 3, second half of the second paragraph (from "Then, we sample a new candidate [...]"): the algorithm 1 is not properly described. Please provide a description in line with algorithm 1 and, above all, describe the rationale behind the choice of line 7 of algorithm 1.
I like it, but it deserves to be better described. It seems to me a sort of "novelty-driven" optimization.(4) A small detail, but requires a clarification. In page 6, learning step paragraph: the weight formula uses fitness in the exponent, while the base rho is in (0,1). Clearly the authors are working with minimization problems and want to increase the weight when fitness improves/decreases. This is ok, but this only happens with their formula when fitnesses are greater than 1. I know this is quite usual in the problems analyzed by the authors, but it may not be so usual in general. Please add a clarification.}

The key idea behind UMM is that the probabilistic Mallows model (MM) used in
UMM does not simply represent the current sample $S$. Instead, the model
assigns higher probability values to the permutations with known (or expected)
minimal fitness values. This is achieved by means of the uBorda ranking
(Section~\ref{sec:uborda}) with $w(\sigma) = \rho^{f(\sigma)}$ and $\rho \in (0,1)$. %, which %  can be seen as replicating each permutation
Intuitively, uBorda behaves as if there were many copies of the best
solutions and few of the bad ones. A property of UMM is that the
computational complexity of working with the original sample (Borda) and with the
transformed sample (uBorda) is the same, i.e., polynomial. The most important steps are the update
(\emph{learning}) of the parameters $\sigma_0$, $\theta$ and $\rho$ (line~\ref{line:uborda} in Algorithm~\ref{alg:umm}) and
the sampling from $MM(\sigma_0,\theta)$ (line~\ref{line:sample}) %(lines~\ref{line:sample}--\ref{line:distance}). % in Algorithm~\ref{alg:umm}).

@@

\subsection{Learning step}
%
At each iteration, we start by fitting a MM to the sample $S$, which implies
estimating $\hat\sigma_0$ and $\theta$. Parameter $\theta$ is updated
deterministically at the sampling step.  The parameter $\sigma_0$, however, is
learned from the current sample $S$ by using the uBorda algorithm with a weight
$w(\sigma)=\rho^{f(\sigma)}$ (Section~\ref{sec:uborda}). The learning rate is
controlled by parameter $0 < \rho < 1$. 
\MANUEL{Don't we have to say something about $f(\sigma) > 1$?}
\EKHINE{es decreciente igual}
In UMM, the value of $\rho$ depends on the set of true fitness evaluations $F$
and it is set at each iteration such that the largest $100r_1$\% of the mass of
the weights is concentrated in the best $100r_2$\% of the solutions in $S$:
%
\MANUEL{Reviewer 3:  Justify the form of the formula used to learn the "learning rate" $\rho$ (equation 4). Why this particular approach was chosen?}
% 
\begin{equation}
%  \begin{split}
    r_1 \cdot \sum_{\tuple{\sigma,f(\sigma)}\in S}\rho^{f(\sigma)} =  \sum_{\tuple{\sigma', f(\sigma')}\in S'}\rho^{f(\sigma')}  %\\
  %    & \text{where}\;|S'| = r_2|S| \land \forall \sigma'\in S', \sigma\in S\setminus S' : f(\sigma')\leq f(\sigma).
%  \end{split}
\end{equation}
%
where $|S'| = r_2|S| \land \forall \tuple{\sigma',f(\sigma')}\in S', \tuple{\sigma,f(\sigma)}\in S\setminus S' : f(\sigma')\leq f(\sigma)$, that is, $S'$ contains the $r_2|S|$ elements of $S$ with the best  fitness values. %The values of $r_1, r_2 \in (0, 1)$ are user-defined and control the balance between exploration and intensification. %: extreme different values bias the sampling towards the best solutions found so far, which may lead to quick stagnation, whereas values (0.5,0.5) make the sampling more uniform, thus slowing down the learning.

@@

%\paragraph{Quality guarantees}

@@

\subsubsection{UMM}
%
UMM uses a budget of evaluations ($\FEmax$) of  $\minit=10$ solutions sampled uniformly.
The other two parameters of UMM are $r_1$ and $r_2$ that determine the learning rate $\rho$ in uBorda, and consequently, determine the
balance between exploration and exploitation. In the experimental section we
will study their impact.

In all
experiments, we consider a maximum budget of $\FEmax=400$ evaluations of the
actual objective function. In a white-box context, state-of-the-art algorithms
for the LOP and PFSP typically evaluate thousands of solutions,
thus, the budget considered here for the black-box context is extremely
limited.

@@

We use the original implementation of GECO.\footnote{\url{https://cran.r-project.org/package=CEGO}} %
Although it is never stated in the original paper, the implementation of CEGO
generates a set of initial solutions of size $\minit=10$ by means of a
max-min-distance sequential design: new solutions are added to the set
sequentially by maximizing the minimum distance to solutions already in the
set. These initial solutions are then evaluated on the true fitness function
and the result is used to build the initial surrogate
model. % \citet{ZaeStoBar2014:ppsn} uses \minit=10
Following the authors of
CEGO~\citep{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn}, we use a GA to optimize
\ekhine{}{hay que actualizar essto??}
the surrogate models with population size of 20, crossover rate of 0.5,
mutation rate of $1/n$, tournament selection of size 2 with probability of 0.9,
interchange mutation (i.e., exchanging two randomly selected elements) and
cycle crossover for permutations. To generate each new solution, CEGO runs the
GA using the model as a surrogate of the true fitness function.  The authors of
CEGO used a budget of $10^4$ surrogate-model evaluations for each run of the
GA and we use the same confuguration parameters.% However, with such a setting, a single run of CEGO with a budget of
%$\FEmax=400$ true fitness evaluations requires \emph{more than a week} of
%computation time, which was infeasible for our study. 
%Due to extremely long
%runtimes of CEGO and constraints in our computing system, we set the budget of
%the GA to $10^3$ and we stopped each run of CEGO after 144 hours of CPU-time.
%Although a higher GA budget could improve the results of CEGO reported here,
%our results are similar (and sometimes better) than the ones reported in the
%original papers~\citep{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn} due to
%running CEGO with $\FEmax=400$ instead of $\FEmax=200$. We found that
%doubling $\FEmax$ produces a smaller increase in runtime than multiplying the
%GA budget by a factor of $10$. Nevertheless, we caution that choosing a
%different trade-off may lead to different conclusions and more work is needed
%to understand CEGO parameters.

\subsubsection{Other settings}
%
For simplicity, we use Kendall's-$\tau$
distance~(Eq.~\ref{eq:kendall_decomp}) in all experiments.
%Nevertheless,
%\citet{ZaeStoFriFisNauBar2014} points out that CEGO performs best on the QAP
%when using Hamming distance and the same may apply to UMM.
We plan to extend UMM to other distance measures in the future.

\subsubsection{Benchmark problems}
%
Experiments with real-world expensive black-box problems would be computationally infeasible. Instead, we consider classical combinatorial optimization problems as black-box optimization problems. For consistency with the rest of the paper, a permutation $\sigma$ denotes a ranking also in this section. Recall that the ordering associated to ranking $\sigma$ is given by its inverse, $\sigma^{-1}$. Thus, we will use ordering $\sigma^{-1}$ to formulate the problems. Later in the experimental section we discuss the effects of not performing this inversion.

The \emph{Linear Ordering Problem (LOP)}, when minimizing as in our case, is defined as
%\footnote{For consistency with the other problems, we minimise the lower triangle instead of maximising the upper triangle.} 
\begin{equation}\label{eq:lop}
  \min_{\sigma \in S_n} f(\sigma) = \sum_{i=1}^{n} \sum_{j=1}^{i-1}  a_{\sigma^{-1}(i), \sigma^{-1}(j)}
\end{equation}
%
where $[a_{i,j}]$ is a matrix of size $n \times n$.

@@

The \emph{Permutation Flowshop Scheduling Problem (PFSP)} is defined by a matrix $[p_{ij}]$ of size $n \times M$ that gives the processing time of a job $i \in [n]$ on a machine $j \in [M]$. All jobs must be processed by all machines in the same order. Given a permutation (ranking) $\sigma$, $C_{i,j}$ denotes the completion time of job $\sigma^{-1}(i)$ at position $i$ on machine $j$, and $C_{\max} = C_{n,M}$  is the completion time of the last job on the last machine, i.e., the makespan. The objective of the PFSP is to find:
\begin{equation}\label{eq:pfsp}
  \begin{split}
    \min_{\sigma \in S_n}\quad& f(\sigma) = C_{n,M}\\
    \text{s.t.}\quad & C_{1,j} = 0 \quad j\in [M], \qquad C_{i,1} = 0\quad i \in [n]\\
    &C_{i,j} = p_{\sigma^{-1}(i),j} + \max\{C_{i-1, j}, C_{i, j-1}\}  \ \  i \in \{2,\dotsc,n\},\\
    &\hspace{14.5em} j \in\{2,\dotsc, M\}
  \end{split}
\end{equation}

@@

\subsubsection{Computing environment}
% Due to the computational effort required, we used several computing systems to
% run experiments.
% All UMM experiments were either on run on Intel Xeon
% \emph{Broadwell} E5-2683v4 CPUs at 2.10\,GHz or Intel Xeon \emph{Haswell} E5-2680v3
% CPUs at 2.50\,GHz, both of them with 32\,GB RAM. 
 %
Experiments were run on Intel Xeon \emph{Ivybridge} E5-2650v2 CPUs at
2.60\,GHz, 64\,GB RAM running CentOS Linux release 7 (Core).
%The operating system was CentOS Linux release 7 (Core).
